MCML with Boundary Surfaces

### 2019-05-29 docs/report/article.tex
Update report

Monte Carlo simulation of light is very accurate, because it simulates single photon packets and records the distribution of a very large number of photons. This implies, that it is a statistical method, using a pseudo random number generator to compute, what can be thought of random walks of photons, visualized in \autoref{trajectory}.

@@

Wang and Jacques~\cite{wang1992monte} published a C implementation of MCML in 1992 along with extensive documentation. The code is purely sequential and because of the inherently large number of photons required, it was worth working on another implementation. Alerstam et al.~\cite{alerstam2008parallel} published a GPU implementation in 2008, which uses the Nvidia CUDA interface. They achieved much faster simulations, but their program runs only on single Nvidia GPUs.

@@

The implementation of MCML presented here can use multiple PCs and GPUs by any vendor for parallelization and is based on OpenCL and MPI. Details are described in the implementation section.

The other contribution is support for non-planar layers. This is done by representing boundaries as radially symmetric heightfields, where the symmetry allows to store only a 1-dimensional array of height samples. Non-symmetric heightfields would only make sense, once photon distributions are also recorded in two dimensions. Additionally, each height value is associated with a spacing, to enable irregular grids. The surface between height samples is defined by linear interpolation. In 3-dimensional space, a surface is therefore a sequence of connected concentric capped cones of finite height.

@@

@@

\subsection{Boundary Intersection}

@@

@@

To obtain a height value at arbitrary floating point coordinates, the heightfield must be interpolated. Therefore it would make sense to store them in textures and use built-in access operators on the GPU. Currently, the program uses raw buffers and does interpolation in code.

@@

@@

\section{Conclusion}

>>>>

Monte Carlo simulation of light is very accurate, because it simulates single photon packets and records the distribution of a large number of photons. This implies, that it is a statistical method, using a pseudo random number generator to compute, what can be thought of random walks of photons, visualized in \autoref{trajectory}.

@@

Wang and Jacques~\cite{wang1992monte} published C source code implementing MCML in 1992 along with extensive documentation. The program is purely sequential and because of the inherently large number of photons required, it was worth working on another implementation.

Alerstam et al.~\cite{alerstam2008parallel} published a GPU implementation in 2008, which uses the Nvidia CUDA interface. The source code, user manual and implementation notes are available as download from the linked website, as of now. They achieved much faster simulations, but their program runs only on single Nvidia GPUs.

@@

The implementation of MCML presented here, that I want to call clustermcml, can use multiple PCs and GPUs by any vendor for parallelization and is based on OpenCL and MPI. Details are described in the implementation section.

The other contribution is support for non-planar layers. This is done by representing boundaries as radially symmetric heightfields. The symmetry allows to store only a 1-dimensional array of height samples. Non-symmetric heightfields would only make sense, once photon distributions are also recorded in two dimensions. Additionally, each height value is associated with a spacing, to enable irregular grids. The surface between height samples is defined by linear interpolation. In 3-dimensional space, a surface is therefore a sequence of connected concentric capped cones of finite height.

@@

When a boundary collision is detected, the photon moves a shorter path than the previously calculated step length. The original MCML implementation saves the remaining step length for the next step, but CUDAMCML authors argue, that the randomized next step length is independent of its history. The implementation presented here therefore also does not save the remaining step length, but ignores it.

@@

\subsection{Boundary Intersection Test}

@@

There is another make target for an executable called singlemcml-windows, which is a non-parallel and CPU-only version of the program for the purpose of easier debugging.

@@

There are minor deviations from original MCML. The following points are very similar to the limitations of CUDAMCML, as can be seen in their manual in section 2.4.

\begin{itemize}
\item The binary output is currently not supported. MCO files are always written as ASCII.
\item Glass layers are not supported. Reflectance will always be reported as diffuse.
\item The maximum number of photons is limited to $2^{32}-1$. Also the number of photons must be higher than the number of GPU threads, which will be printed on program startup.
\item Input data is not checked for sanity.
\item All calculations are preformed with 32-bit floating point precision. This is also the case for CUDAMCML and was shown to have only slight effect.
\end{itemize}

The implementation currently leaves much room for performance optimization. Important aspects are listed here.

\begin{itemize}
\item The array of photon states is iterated sequentially after each GPU run, to check for photons to restart. This operation can be parallelized or integrated into the already parallel kernel. CUDAMCML restarts finished photons immediately from the kernel. This also avoids inactive threads, but introduces the cost of another atomic counter to keep track of total finished photons.

\item GPUs have global and local memory spaces, where access to local memory is usually faster. Therefore, CUDAMCML accumulates weights in local variable before performing an atomic add on a global output array. The implementation presented here performs the atomic add every time, when a weight is stored.

\item To read a height value from a boundary at arbitrary floating point coordinates, the heightfield must be interpolated. Therefore it would be efficient to store the heightfields in textures and use built-in access operators on the GPU. Currently, the program uses raw buffers and does interpolation in code.
\end{itemize}

@@

\autoref{verification1} compares the diffuse reflectance profiles of clustermcml and CUDAMCML using 100 million photons and one layer of thickness $dl=0.1cm$. The absorption coefficient is $\mu_a=0.1$, the scattering coefficient $\mu_s=1$ and scattering anisotropy $g=0.5$. The refractive indices are $n=1.34$ inside the layer and $1$ outside. The detection window has a size of $1cm$ and a resolution of $0.005cm$. Note that a logarithmic scale is used for a quantity, that is reflectance multiplied by radius. The plots indicate correct results despite different amounts of noise.

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification1.pdf}
	\caption{Comparison of the diffuse reflectance profiles obtained with the implementation presented in this report, clustermcml, and the previous implementation CUDAMCML as reference.}
	\label{verification1}
\end{figure}

@@

This section shows novel radiance profiles that arise from simulations with different boundary surface shapes.

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result1.pdf}
	\caption{Comparison of diffuse reflectance profiles of 7 different surfaces.}
	\label{result1}
\end{figure}

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result2.pdf}
	\caption{Comparison of diffuse reflectance profiles of flat surface with 4 variations of curved surfaces.}
	\label{result2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/result3.pdf}
\caption{Comparison of diffuse reflectance profiles of flat surface with 2 variation of noise surfaces.}
\label{result3}
\end{figure}

\section{Conclusion and Future Work}
